No need to know: astrophysics-free gravitational-wave cosmology

### 2024-08-22 src/tex/ms.tex
[showyourwork] overleaf sync

We find that our non-parametric model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ during \ac{O5}, when the detectors will reach their design sensitivity. 
% \new{This is to be compared with a $\variable{output/PLPh0percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ with the correct parametric model, which does not represent a statistically-significant difference.}
We highlight measurements of \Ho{} within a $\Lambda$CDM universe in order to benchmark the accuracy and precision of our non-parametric method, as well as explore the role of spectral sirens in elucidating the Hubble tension.
However, a primary utility of spectral siren measurements will be in constraining $H(z)$ under different cosmological models, and at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{You:2020wju, Ezquiaga:2020tns, ezquiaga_spectral_2022, Chen:2024gdn}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
%When applied to data from next-generation detectors, this method will measure \new{cosmological parameters that govern $H(z)$.}

@@

To obtain a quantitative measure of the systematic bias introduced by mismodeling the mass distribution, we repeat the parametric analyses with 50 separate simulated catalogs of $\sim1,000$ events each.
We find that the \bpl{} model produces an over- or under-estimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, 
and the \plp{} model reaches the same level of bias only $\variable{output/PLP_bias_percent.txt}\%$ 
of the time, \new{meaning that the \bpl{} model produces a bias more than three times as often as the \plp{} model.}
\new{Additionally, we show in Appendix~\ref{ap:parametric bias} that the \bpl{} model typically overestimates \Ho{} whereas the \plp{} model produces a roughly equal number of over- and underestimates of \Ho{}.
This implies a systematic bias in the incorrect but not the correct parametric model.}

Collectively, these results indicate that prior knowledge of the shape of the mass distribution is not required to perform an unbiased spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.

@@

\bibliography{bib}
\appendix
\section{Details of data simulation}
\label{ap:data generation}
The exact form of the injected population is
\begin{equation}
\label{eq:underlying population}
    \frac{\diff N}{\diff m_1 \diff z} \propto p(m_1|\bar{\Lambda}_m) p(z|\bar{\Lambda}_z, H_0, \Omega_M),
\end{equation}
where 
\begin{align}
    p(m_1,m_2|\bar{\Lambda}_m) &\propto \mathcal{S}(m_{\min},m_{\max})
    \left( f_{\text{peak}}e^{-\frac{1}{2}(\frac{m_1-\mu}{\sigma})^2}\mathcal{N}_{\text{g}} +
    (1-f_{\text{peak}})m_1^{\alpha}\mathcal{N}_{\text{pl}} \right) ,
\label{eq:underlying mass dist}
\end{align}
\begin{equation}
    p(z|H_0, \Omega_M) \propto \frac{\diff V_C}{\diff z} \frac{1}{1+z} \frac{(1+z)^\alpha_z}{1+\left(\frac{1+z}{1+z_p}\right)^{\alpha_z+\beta_z}}.
    \label{eq:underlying redshift dist}
\end{equation}
Here, $V_C(H_0, \Omega_M)$ is the comoving volume for given cosmological parameters \Ho{} and \Omm{}, and $\bar{\Lambda}_m = \{\alpha, m_{\min}, m_{\max}, \mu, \sigma, f_{\text{peak}}\}$ are the (hyper-)parameters describing the power law in primary mass, minimum and maximum black hole mass, Gaussian peak location and width, and fraction of events in the Gaussian peak, respectively.
$\mathcal{S}$ is a smoothing function at low and high masses, and $\mathcal{N}_{\text{pl}}$ and $\mathcal{N}_{\text{g}}$ are the normalizations between  $m_{\min}$ and $m_{\max}$ for the power law component and truncated Gaussian component, respectively.
The smoothing creates support for masses below $m_{\min}$ and above $m_{\max}$.
$\bar{\Lambda}_z = \{z_p,\alpha_z,\beta_z\}$ are the parameters governing the peak of the redshift distribution, low-$z$ power law slope, and high-$z$ power law slope.
When generating the simulated events, we have fixed $\alpha=-2.7$, $m_{\max}=78\Msun$, $m_{\min}=10~\Msun$, $\mu=30~\Msun$, $\sigma=7.0~\Msun$, $f_{\text{peak}}=0.05$, $z_p=2.4$, $\alpha_z=1.0$, and $\beta_z=3.4$.
These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{abbott_population_2023}.
For simplicity, we additionally assume \new{a uniform mass ratio distribution.}
Thus, $\theta = \{m_1,z\}$.
We consistently apply these assumptions to the data generation process and the population inference.
We do not fit for or simulate spins, as they do not redshift and hence do not carry additional cosmological information, and additionally \new{fix the distribution of mass ratios.}

@@

\new{We use the same set of simulated events for both analyses.
These are generated in the same way as described in Section~\ref{sec:data generation}. The results are shown in Figure~\ref{fig:mass ratio}.
We find the posteriors on \Ho{} to be similar between the two cases.
Additionally, the recovered mass spectra are nearly identical.
We therefore conclude that fitting for mass ratios does not significantly impact our main conclusions.
}
\begin{figure*}
    \includegraphics[width=\textwidth]{figures/q_comparison.pdf}
    \caption{Comparison of a parametric spectral siren analysis performed while fitting for the distributions of primary and secondary masses (magenta solid) and primary masses only (green dashed).
    The posterior on \Ho{} is relatively unchanged between the two, and only minor changes are observed in the mass spectrum.}
    \label{fig:mass ratio}
    \script{fit_q_comparison.py}
\end{figure*}

@@

\section{Biases observed in parametric models}
\label{ap:parametric bias}
\new{Figure~\ref{fig:parametric bias} shows recovered posteriors on \Ho{} as inferred from 50 mock catalogs using both the \bpl{} and \plp{} parametric models.
We find that the \plp{} model recovers \Ho{} posteriors that are largely symmetric about the true value, whereas the \bpl{} model typically finds more support for larger \Ho{} values.
We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.}

\new{Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.
Therefore, while we are able to identify the shortcomings of \bpl{} when \plp{} describes the true mass distribution, it is not possible in practice to know which direction an incorrect parametric model will bias \Ho{}.
Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.}

>>>>

We find that our non-parametric model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ during \ac{O5}, when the detectors will reach their design sensitivity. 
% \new{This is to be compared with a $\variable{output/PLPh0percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ with the correct parametric model, which does not represent a statistically-significant difference.}
We highlight measurements of \Ho{} within a $\Lambda$CDM universe in order to benchmark the accuracy and precision of our non-parametric method, as well as explore the role of spectral sirens in elucidating the Hubble tension.
However, a primary utility of spectral siren measurements will be in constraining $H(z)$ under different cosmological models, and at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{You:2020wju, Ezquiaga:2020tns, ezquiaga_spectral_2022, Chen:2024gdn}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
%When applied to data from next-generation detectors, this method will measure cosmological parameters that govern $H(z)$.

@@

To obtain a quantitative measure of the systematic bias introduced by mis-modeling the mass distribution, we repeat the parametric analyses with 50 separate simulated catalogs of $\sim1,000$ events each.
We find that the \bpl{} model produces an over- or under-estimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, 
and the \plp{} model reaches the same level of bias only $\variable{output/PLP_bias_percent.txt}\%$ 
of the time, \new{meaning that the \bpl{} model produces a bias more than three times as often as the \plp{} model.}
\new{Additionally, we show in Appendix~\ref{ap:parametric bias} that the \bpl{} model typically overestimates \Ho{} whereas the \plp{} model produces a roughly equal number of over- and underestimates of \Ho{}.
This demonstrates that mis-modeling the mass distribution can introduce statistically significant systematic biases into measurements of cosmological parameters.}

Collectively, our results indicate that \emph{prior knowledge of the shape of the mass distribution is not required to perform an unbiased spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made}.

@@

\bibliography{bib}
\appendix
\section{Details of data simulation}
\label{ap:data generation}
The exact form of the injected population is
\begin{equation}
\label{eq:underlying population}
    \frac{\diff N}{\diff m_1 \diff z} \propto p(m_1|\bar{\Lambda}_m) p(z|\bar{\Lambda}_z, H_0, \Omega_M),
\end{equation}
where 
\begin{align}
    p(m_1,m_2|\bar{\Lambda}_m) &\propto \mathcal{S}(m_{\min},m_{\max})
    \left( f_{\text{peak}}e^{-\frac{1}{2}(\frac{m_1-\mu}{\sigma})^2}\mathcal{N}_{\text{g}} +
    (1-f_{\text{peak}})m_1^{\alpha}\mathcal{N}_{\text{pl}} \right) ,
\label{eq:underlying mass dist}
\end{align}
\begin{equation}
    p(z|H_0, \Omega_M) \propto \frac{\diff V_C}{\diff z} \frac{1}{1+z} \frac{(1+z)^\alpha_z}{1+\left(\frac{1+z}{1+z_p}\right)^{\alpha_z+\beta_z}}.
    \label{eq:underlying redshift dist}
\end{equation}
Here, $V_C(H_0, \Omega_M)$ is the comoving volume for given cosmological parameters \Ho{} and \Omm{}, and $\bar{\Lambda}_m = \{\alpha, m_{\min}, m_{\max}, \mu, \sigma, f_{\text{peak}}\}$ are the (hyper-)parameters describing the power law in primary mass, minimum and maximum black hole mass, Gaussian peak location and width, and fraction of events in the Gaussian peak, respectively.
$\mathcal{S}$ is a smoothing function at low and high masses, and $\mathcal{N}_{\text{pl}}$ and $\mathcal{N}_{\text{g}}$ are the normalizations between  $m_{\min}$ and $m_{\max}$ for the power law component and truncated Gaussian component, respectively.
The smoothing creates support for masses below $m_{\min}$ and above $m_{\max}$.
$\bar{\Lambda}_z = \{z_p,\alpha_z,\beta_z\}$ are the parameters governing the peak of the redshift distribution, low-$z$ power law slope, and high-$z$ power law slope.
When generating the simulated events, we have fixed $\alpha=-2.7$, $m_{\max}=78\Msun$, $m_{\min}=10~\Msun$, $\mu=30~\Msun$, $\sigma=7.0~\Msun$, $f_{\text{peak}}=0.05$, $z_p=2.4$, $\alpha_z=1.0$, and $\beta_z=3.4$.
These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{abbott_population_2023}.
For simplicity, we assume \new{a uniform mass ratio distribution.}
Thus, $\theta = \{m_1,z\}$.
We consistently apply these assumptions to the data generation process and the population inference.
We do not fit for or simulate spins, as they do not redshift and hence do not carry additional cosmological information, and \new{we fix the distribution of mass ratios.}

@@

\new{We use the same set of simulated events for both analyses.
These are generated in the same way as described in Section~\ref{sec:data generation}.
The inference results are shown in Figure~\ref{fig:mass ratio}.
We find the posteriors on \Ho{} to be similar between the two cases.
Additionally, the recovered mass spectra are nearly identical.
We therefore conclude that fitting for mass ratios does not significantly impact our main conclusions.
}
\begin{figure*}
    \includegraphics[width=\textwidth]{figures/q_comparison.pdf}
    \caption{Comparison of a parametric spectral siren analysis performed while fitting for the distributions of primary and secondary masses (magenta solid) and primary masses only (green dashed).
    The posterior on \Ho{} is relatively unchanged between the two, and only minor differences are observed in the mass spectra.}
    \label{fig:mass ratio}
    \script{fit_q_comparison.py}
\end{figure*}

@@

\section{Biases induced by the \bpl{} model}
\label{ap:parametric bias}
\new{Figure~\ref{fig:parametric bias} shows recovered posteriors on \Ho{} as inferred from 50 mock catalogs using both the \bpl{} and \plp{} parametric models.
We find that the \plp{} model recovers \Ho{} posteriors that are largely symmetric about the true value, whereas the \bpl{} model typically finds more support for larger \Ho{} values.
We note that \citep{pierra_study_2023} report an underestimate of \Ho{} when fitting their Broken Power Law model to a single dataset generated with \plp, but the prior bounds on the maximum mass parameter of their model do not extend below the injected value of that parameter, which in turn does 
not allow for the possibility of an overestimated \Ho{}.
It is therefore not possible to directly compare the direction of our observed \Ho{} offset with theirs, despite the similarities bewteen our chosen population models.}

\new{Regardless, this systematic offset demonstrates that incorrect parametric models unfortunately induce a systematic bias in cosmological inference, not an increase in statistical uncertainty. 
Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.
Therefore, while we are able to identify the shortcomings of \bpl{} when \plp{} describes the true mass distribution, it is not possible in practice to know which direction an incorrect parametric model will bias \Ho{}.
Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.}
